Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-12-22 threshold/draft.Rnw
Updated draft

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$,
compute $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
threshold = 0.95

# Register a four-core multicore parallel backend
registerDoMC(4)

@@

# Fast brute force full correlation:
m2  = sum(gc()[,2])
t2  = proc.time()
mu = colMeans(A)
s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
cs = scale(A, center=mu, scale=s)  # scale and center the data
cx = crossprod(cs)                 # full correlation matrix
cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

@@

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  library(doMC)
  system("wget http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/ACC/20150821/gdac.broadinstitute.org_ACC.Merge_methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.Level_3.2015082100.0.0.tar.gz")
  cmd = "cat ACC.methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.data.txt  | tail -n +%s | cut -d '       ' -f 2,6,10,14,18,22,26,30,34,38,42,46,50,54,58,62,66,70,74,78,82,86,90,94,98,102,106,110,114,118,122,126,130,134,138,142,146,150,154,158,162,166,170,174,178,182,186,190,194,198,202,206,210,214,218,222,226,230,234,238,242,246,250,254,258,262,266,270,274,278,282,286,290,294,298,302,306,310,314,318"
  N = 485579  # total number of lines
  h = 2       # total header lines
  # Read the first header line with sample IDs:
  f = pipe(cmd, open="r")
  id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
  close(f)
  # Read the rest of the file in parallel. I prefer doMC/foreach over mclapply for this because foreach can combine results incrementally.
  registerDoMC(4)  # four cores on our test machine
  block = floor((N - h)/cores)
  acc = foreach(j=1:cores, .combine=rbind) %dopar%
  {
    skip = block * (j - 1) + h + 1
    nrows = ifelse(j == cores, -1, block)
    f = pipe(sprintf(cmd, skip), open="r")
    on.exit(close(f))
    read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=nrows)
  }
  acc = t(acc)
  # OK, the example data are parsed and in acc. proceed as in the smaller example.
  threshold = 0.99
  # Compute using tcor():
  m1  = sum(gc()[,2])  # Memory in megabytes currently in use
  t1  = proc.time()
  tx  = tcor(acc, t=threshold, p=10)
  t1  = (proc.time() - t1)[3] # Time in seconds
  m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test

>>>>

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$,
compute $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
One can cheaply estimate whether there are ``too many'' candidate pairs for a
given dimension $p$ in step 5 by evaluating step 4 for only adjacent pairs of
vectors, $\tilde{i}=1$.

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
threshold = 0.95

# Test system values:
t1=3.13
t2=1.75
m1=59.60
m2=1075.10

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
# Register a four-core multicore parallel backend
registerDoMC(4)

@@

# Fast brute force full correlation:
m2  = sum(gc()[,2])
t2  = proc.time()
mu = colMeans(A)
s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
cs = scale(A, center=mu, scale=s)  # scale and center the data
cx = crossprod(cs)                 # full correlation matrix
cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2
}
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

@@

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  library(doMC)
  system("wget http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/ACC/20150821/gdac.broadinstitute.org_ACC.Merge_methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.Level_3.2015082100.0.0.tar.gz")
  cmd = "cat ACC.methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.data.txt  | tail -n +%s | cut -d '       ' -f 2,6,10,14,18,22,26,30,34,38,42,46,50,54,58,62,66,70,74,78,82,86,90,94,98,102,106,110,114,118,122,126,130,134,138,142,146,150,154,158,162,166,170,174,178,182,186,190,194,198,202,206,210,214,218,222,226,230,234,238,242,246,250,254,258,262,266,270,274,278,282,286,290,294,298,302,306,310,314,318"
  N = 485579  # total number of lines
  h = 2       # total header lines
  # Read the first header line with sample IDs:
  f = pipe(cmd, open="r")
  id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
  close(f)
  # Read the rest of the file in parallel. I prefer doMC/foreach over mclapply for this because foreach can combine results incrementally.
  registerDoMC(4)  # four cores on our test machine
  block = floor((N - h)/cores)
  acc = foreach(j=1:cores, .combine=rbind) %dopar%
  {
    skip = block * (j - 1) + h + 1
    nrows = ifelse(j == cores, -1, block)
    f = pipe(sprintf(cmd, skip), open="r")
    on.exit(close(f))
    read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=nrows)
  }
  acc = t(acc)

  # remove columns containing missing values, all zero
  n = ncol(acc)
  acc = acc[, colMeans(acc) > 0]
  cat("Removed", n - ncol(acc), "columns of all zeros.\n")
  n = ncol(acc)
  acc = acc[, apply(acc,2,function(x) !any(is.na(x)))]
  cat("Removed", n - ncol(acc), "columns with missing values.\n")
  cat("Data matrix size ", nrow(acc), "x", ncol(acc), "\n")

  # OK, the example data are parsed and in acc. proceed as in the smaller example.
  threshold = 0.99
  # Compute using tcor():
  m1  = sum(gc()[,2])  # Memory in megabytes currently in use
  t1  = proc.time()
  tx  = tcor(acc, t=threshold, p=10)
  t1  = (proc.time() - t1)[3] # Time in seconds
  m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test

### 2015-12-22 threshold/draft.tex
Updated draft

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
\begin{Schunk}
\begin{Sinput}
> library(tcor)
> library(doMC)     # local parallel computation
> library(biclust)  # for the BicatYeast data
> data(EisenYeast)  # in gene by sample orientation
> A = t(EisenYeast) # to correlate genes to genes
> threshold = 0.95
> # Register a four-core multicore parallel backend
> registerDoMC(4)
> # Compute using tcor():
> m1  = sum(gc()[,2])  # Memory in megabytes currently in use
> t1  = proc.time()
> tx  = tcor(A, t=threshold, p=10)
> t1  = (proc.time() - t1)[3] # Time in seconds
> m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
> # Fast brute force full correlation:
> m2  = sum(gc()[,2])
> t2  = proc.time()
> mu = colMeans(A)
> s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
> cs = scale(A, center=mu, scale=s)  # scale and center the data
> cx = crossprod(cs)                 # full correlation matrix
> cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
> pairs = which(cx >= threshold, arr.ind=TRUE)
> t2  = (proc.time() - t2)[3]
> m2  = sum(gc()[,6]) - m2
\end{Sinput}
\end{Schunk}
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

% latex table generated in R 3.2.3 by xtable 1.8-0 package
% Sun Dec 20 09:52:31 2015
\begin{table}[ht]
\centering
{\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.19 & 59.60 \\ 
  brute force & 1.74 & 1075.10 \\ 
   \hline
\end{tabular}
}
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.3 by xtable 1.8-0 package
% Sun Dec 20 09:52:31 2015
\begin{table}[ht]
\centering
{\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
}
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

>>>>

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95. The choice of projection dimension $p$ affects
the efficiency of the algorithm. For $p=2$ the number of candidate vector pairs
generated in step 4 of Algorithm \ref{prune} is over 3 million. When $p$ is increased
to $10$, the number of candidate vector pairs drops to less than 13,000.
The reference \proglang{R} implementation includes options for
tuning $p$ and relies on the restart ability of the IRLBA to pick up where it left off and
inexpensively extend an existing truncated SVD. The following example uses $p=10$.
\begin{small}
\begin{Schunk}
\begin{Sinput}
> threshold = 0.95
> # Test system values:
> t1=3.13
> t2=1.75
> m1=59.60
> m2=1075.10
> if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
+ {
+ library(tcor)
+ library(doMC)     # local parallel computation
+ library(biclust)  # for the BicatYeast data
+ data(EisenYeast)  # in gene by sample orientation
+ A = t(EisenYeast) # to correlate genes to genes
+ # Register a four-core multicore parallel backend
+ registerDoMC(4)
+ 
+ # Compute using tcor():
+ m1  = sum(gc()[,2])  # Memory in megabytes currently in use
+ t1  = proc.time()
+ tx  = tcor(A, t=threshold, p=10)
+ t1  = (proc.time() - t1)[3] # Time in seconds
+ m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
+ 
+ # Fast brute force full correlation:
+ m2  = sum(gc()[,2])
+ t2  = proc.time()
+ mu = colMeans(A)
+ s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
+ cs = scale(A, center=mu, scale=s)  # scale and center the data
+ cx = crossprod(cs)                 # full correlation matrix
+ cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
+ pairs = which(cx >= threshold, arr.ind=TRUE)
+ t2  = (proc.time() - t2)[3]
+ m2  = sum(gc()[,6]) - m2
+ }
\end{Sinput}
\end{Schunk}
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

% latex table generated in R 3.2.3 by xtable 1.8-0 package
% Mon Dec 21 19:50:10 2015
\begin{table}[ht]
\centering
{\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.13 & 59.60 \\ 
  brute force & 1.75 & 1075.10 \\ 
   \hline
\end{tabular}
}
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.3 by xtable 1.8-0 package
% Mon Dec 21 19:50:10 2015
\begin{table}[ht]
\centering
{\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
}
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.
